\Large Generative Dynamics of Supreme Court Citations

### 2018-04-30 Tex/paper.tex
finished estimation

Though the likelihood function of the c-TERGM may appear quite different from that of statistical models conventionally used in political science, analysis of the conditional probability of a single citation from case $i$ to case $j$ reveals that we can interpret the parameters similar to logistic regression coefficients. $$ P(C_{ij,t} = 1 | C_{-ij,t}, C_{ < t}) = \frac{\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right]}{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right] + \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 0) } \right]}, $$ $$ \text{~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~} = \frac{1}{ 1 + \exp \left[ - {\bm{\theta}'\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)} \right]}, $$
where $C_{ij,t} = 1$ indicates that case $i$ cites case $j$, $C_{ij,t} = 0$ indicates that case $i$ does not cite case $j$, $C_{-ij,t}$ is the observed elements of $C_{t}$ except $C_{ij,t}$, and $\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)$ is the change in $\bm{h}(C_{t},C_{<t})$ that results from toggling $C_{ij,t} = 0$ to $C_{ij,t} = 1$. This re-arrangement illustrates that the parameters can be interpreted in terms of the change in the log odds of a citation from $i$ to $j$ given a one-unit increase in the corresponding element of $\bm{h}$, conditional on the other citations observed in the network. For example, if the value of $\theta$ corresponding to an element of $\bm{h}$ that counts the number of mutual edges in the network is 0.5, then the log odds of observing $C_{ij,t} = 1$ increase by 0.5 if case $i$ is cited by case $j$ (as compared to the configuration in which case $j$ does not cite case $i$).

@@

The normalizing constant in Equation 1 is intractable. For example, in the simple case of adding three cases to a network in which six cases already exists---like that depicted in Figure \ref{fig:ctergm}---there are 16,777,216 unique configurations of $C_t$ that could be observed. The typical Supreme Court term involves adding hundreds of cases to a network that

\indent The normalizing constant in the denominator of the probability function (i.e., Equation \ref{ercm}) is computationally intractable. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with ERGM family models. Methods of Monte Carlo approximation are well established \citep{hunter2006inference,van2009framework,hummel2012improving}, but with almost $10,000$ vertices, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. To estimate the parameters of the c-TERGM, we rely on two features of the model and data that facilitate the application of an approximate inference method that is faster than Monte Carlo approximation---maximum pseudolikelihood. First, our sample size is very large. Second, under the c-TERGM we assume that citations are formed in each time period conditional upon the citations that exist in the past.  We estimate the c-TERGM via bootstrapped pseudolikelihood \citep{desmarais2012statistical,desmarais2010consistent}.

Before we transition to the empirical analysis, we should note that there is nothing about the c-TERGM (or any ERGM family model, for that matter) that permits the identification of causal effects of one tie on the formation of another. For example, if we find a positive popularity effect, that could be driven by one of two (or some combination of both) dynamics. First, it could be the case that a large volume of citations to a case in past opinions draws the attention of the justices to the ``popular'' case. Second, it could be that we have omitted a case-level variable that is strongly associated with the number of citations accumulated by a case, and the number of citations in past cases is operating as a proxy for this unmeasured case characteristic. We cannot statistically differentiate between these two dynamics. However, in both of these conditions, and analogous conditions related to other dependence properties, citations (or non-citations) to a case are not independent conditional on the other variables, and the c-TERGM can be specified to account for the dependence in the data.

>>>>

Though the likelihood function of the c-TERGM may appear quite different from that of statistical models conventionally used in political science, analysis of the conditional probability of a single citation from case $i$ to case $j$ reveals that we can interpret the parameters similar to logistic regression coefficients. $$ P(C_{ij,t} = 1 | C_{-ij,t}, C_{ < t}) = \frac{\exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right]}{ \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) } \right] + \exp \left[ {\bm{\theta}'\bm{h}(C_{t},C_{<t}| C_{ij,t} = 0) } \right]}, $$ $$ \text{~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~} = \frac{1}{ 1 + \exp \left[ - {\bm{\theta}'\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)} \right]}, $$
where $C_{ij,t} = 1$ indicates that case $i$ cites case $j$, $C_{ij,t} = 0$ indicates that case $i$ does not cite case $j$, $C_{-ij,t}$ is the observed elements of $C_{t}$ except $C_{ij,t}$, and $\left(\bm{h}(C_{t},C_{<t}| C_{ij,t} = 1) - \bm{h}(C_{t},C_{<t}| C_{ij,t} = 0)\right)$ is the change in $\bm{h}(C_{t},C_{<t})$ that results from toggling $C_{ij,t} = 0$ to $C_{ij,t} = 1$. This re-arrangement illustrates that the parameters can be interpreted in terms of the change in the log odds of a citation from $i$ to $j$ given a one-unit increase in the corresponding element of $\bm{h}$, conditional on the other citations observed in the network. For example, if the value of $\theta$ corresponding to an element of $\bm{h}$ that counts the number of mutual edges in the network is 0.5, then the log odds of observing $C_{ij,t} = 1$ increase by 0.5 if case $i$ is cited by case $j$ (as compared to the configuration in which case $j$ does not cite case $i$). The logit form conditional probability is not unique to c-TERGM. It is well known for the ERGM family \citep{goodreau2009birds}.

We should note that there is nothing about the c-TERGM (or any ERGM family model, for that matter) that permits the identification of causal effects of one tie on the formation of another. For example, if we find a positive popularity effect, the term for that dynamic would indicate that there is an increase in the log odds that case $i$ cites case $j$ with each other case that has cited case $j$.  That finding could be driven by one of two (or some combination of both) dynamics. First, it could be the case that a large volume of citations to a case in past opinions draws the attention of the justices to the ``popular'' case---as in the Baker v. Carr passage cited above. Second, it could be that we have omitted a case-level variable that is strongly associated with the number of citations accumulated by a case, and the number of citations in past cases is operating as a proxy for this unmeasured case characteristic. We cannot statistically differentiate between these two dynamics. However, in both of these conditions, and analogous conditions related to other dependence properties, citations (or non-citations) to a case are not independent conditional on the other variables, and the c-TERGM can be specified to account for the dependence in the data.

@@

The normalizing constant in Equation 1 is intractable. For example, in the simple case of adding three cases to a network in which six cases already exists---like that depicted in Figure \ref{fig:ctergm}---there are 16,777,216 unique configurations of $C_t$ that could be observed. The typical Supreme Court term involves adding hundreds of cases to a network that already includes thousands of previous cases. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with the c-TERGM. Methods of Monte Carlo approximation are well established in the ERGM literature, \citep{hunter2006inference,van2009framework,hummel2012improving}, but with a number of nodes on the order of $10,000$, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. However, because the network is large, we can rely on a fast estimation method---maximum pseudo-likelihood estimation (MPLE), which is consistent for discrete exponential family models \citep{nguyen2017near}, and has long been used with ERGM \citep{strauss1990pseudolikelihood}. The MPLE is defined as maximizing the log product of the conditional probability of each citation (and non-citation), conditional on the other elements of the observed citation network. The joint probability of all citations is replaced by the product over conditional probabilities, which, as we demonstrated above, assume a logit form.

One challenge that arises with the use of MPLE is that the usual estimate of the asymptotic covariance matrix (i.e., inverting the negated Hessian of the objective function at the maximum), results in standard errors that are biased downward. This leads to a high Type-1 error rate in hypothesis testing with MPLE if conventional standard errors are used \citep{van2009framework}. A form of non-parametric cluster bootstrap results in confidence intervals with correct coverage probability with MPLE \citep{desmarais2012statistical,desmarais2010consistent}. In the case of c-TERGM, the clusters are defined by $t$---the time points in which the cases are decided, and $T$ time-specific citation matrices $C_{t}$ are re-sampled with replacement to form a non-parametric cluster bootstrap sample. Confidence intervals are defined by the quantiles in the empirical distributions of estimates that result from re-estimating the MPLE on non-parametric cluster bootstrap samples.
